---
title: "Decades Playlist Analysis"
author: "Hayley Arader"
date: "April 7, 2019"
output: html_document
---

Spotify has a really accessible API!! In this script I use [spotifyr](https://www.rcharlie.com/spotifyr/), which integrates with the Spotify Web API using R, to analyze how songs in Spotify's Decades Playlists differ from each other.

In this analysis, I'll be using metrics on the songs in each decades playlist as proxies for understanding what characterized popular music in that decade. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

################################################################################
## Setup #######################################################################
################################################################################

library(spotifyr)
library(tidyverse)
library(stringr)
library(ggplot2)

# I've stored these in my environment
client_id <- Sys.getenv("SPOTIFY_CLIENT_ID")
client_secret <- Sys.getenv("SPOTIFY_CLIENT_SECRET")

# Authenticate!
access_token <- get_spotify_access_token(client_id = client_id,
                                         client_secret = client_secret)

################################################################################
## Plotting Utils ##############################################################
################################################################################

# Set a theme for plots :)
theme_hayley <- function(...) {

  # first we'll want to update all text
  ggplot2::update_geom_defaults(
    "text",
    list(family = "Raleway", size = 4, fontface = "plain", color = "black",
      hjust = .5, vjust = .5))

  # return theme
  theme_bw() +
    theme(plot.background = ggplot2::element_rect(fill = "white"),
          panel.background = ggplot2::element_rect(fill = "white"),
          panel.border = ggplot2::element_blank(),
          text = element_text(family = "Raleway"))

}

# overwrite ggplot
ggplot <- function(...) ggplot2::ggplot(...) + theme_hayley()

# Function to find music
find_tracks <- function(playlist_data, playlist, metric, n = 10, desc = TRUE) {
  df <- playlist_data[[playlist]] %>% 
    arrange_(.dots = paste0("desc(", metric, ")")) 
  if (!desc) df <- df %>% arrange_(.dots = metric)
  new <- df %>%
    pull(id) %>% 
    .[1:n] %>%
    get_tracks() %>%
    left_join(df %>% select_(.dots = c("id", metric)))
  new$artist <- vapply(new$artists, function(x) x$name[1], character(1))
  
  select_expr <- c("name", "artist", metric)
  names(select_expr) <- str_to_title(select_expr)
  
  new %>% select_(.dots = select_expr)
}
```

## Accessing the data

We can search Spotify for the playlists we're looking for, and filter to the playlist owner. In this case, we're looking for Spotify's Decades playlists, which all begin with `All Out`, and are owned by `Spotify`.
```{r search}
# Spotify has decades playlists, grab them here
playlists <- search_spotify("All Out", type = c("playlist")) %>%
  filter(owner.display_name == "Spotify", str_detect(name, "All Out"))
```

Now we'll read in data about each song in each playlist
````{r read}
# Now read in information about the songs in each playlist
playlist_data <- lapply(seq_along(playlists$id), function(i) {
  get_playlist_tracks(playlists$id[i]) %>%
    pull(track.id) %>%
    get_track_audio_features() %>%
    mutate(name = playlists$name[i])
})
names(playlist_data) <- playlists$name
```

```{r plot, include = FALSE}
################################################################################
## Plotting Functions ##########################################################
################################################################################

# Grab the years of the playlists
years <- str_sub(seq(from = 1950, to = 2010, by = 10), 3, 4)

# Now grab the different metrics spotify provides us with
metrics <- colnames(playlist_data[[1]])[1:11]

# A function to wrangle the playlist data 
# Assumes `playlist_data` is a list of data
# @param data list of data, expects `playlist_data`
# @param metrics_to_plot character of which metrics we're interested in plotting
wrangle_data <- function(data, metrics_to_plot) {
  data %>%
    bind_rows() %>%
    # make the data long
    tidyr::gather(metric, stat, metrics) %>%
    group_by(name, metric) %>%
    # Find stats
    summarize(mean = mean(stat), sd = sd(stat)) %>%
    # let's order by year
    ungroup() %>%
    filter(metric %in% metrics_to_plot) %>%
    mutate(name = factor(name, levels = paste0("All Out ", years, "s")),
           metric = str_to_title(metric))
}

# Function to plot the wrangled data
# @param data dataframe the wrangled data from `wrangle_data`
plot_data <- function(data) {
 data %>%
   ggplot(aes(x = name, color = mean)) +
    # plot error bars
    geom_segment(aes(y = mean - sd, yend = mean + sd, xend = name),
                 color = "grey50", alpha = .5, size = 3) +
    geom_point(aes(y = mean), size = 7) +
    labs(x = "Decade", y = "Mean Value in Playlist") +
    guides(color = F) +
    expand_limits(x = c(-1, 1)) +
    scale_x_discrete(label = function(x) {
      paste0("'", stringr::str_remove(x, "All Out "))
    }) +
    theme(axis.text = element_text(size = 12),
          strip.text = element_text(face = "bold")) +
    facet_wrap(~metric, scales = "free_y")
}

# Put functions together
# @param data list of data, expects `playlist_data`
# @param metrics_to_plot character of which metrics we're interested in plotting
plot_metrics <- function(data, metrics_to_plot) {
  data %>%
    wrangle_data(metrics_to_plot) %>%
    plot_data()
}

# Function to add an annotation explaining what's happening
# in the group
# @param p plot to add the metric to
# @param metric character the name of the metric to add the annotation
add_annotation <- function(p, metric) {
  # dynamically find a yvalue that will make sense
  yval <- (max(p$data$mean) - min(p$data$mean)) / 6
  ymean <- mean(p$data$mean)
  p +
    annotate("label", x = 0, y = ymean,
             label = paste0("Music with\nHigher Avg\n", str_to_title(metric)),
             label.size = 0, family = "Raleway", lineheight = .8,
             fontface = "bold", size = 5) +
    annotate("segment", x = 0, y = ymean + yval, yend = ymean + 2 * yval, xend = 0,
             arrow = arrow(length = grid::unit(.25, "cm")), size = 1)
}

# Function to plot just one metric
# @param data list of data, expects `playlist_data`
# @param metric character the name of the metric to plot
plot_metric <- function(data, metric) {
  plot_metrics(playlist_data, metric) %>%
    add_annotation(metric)
}
```

## Plotting the data

Let's look at all of the metrics, and how they're different across time. 

### Metrics of Interest
These are the metrics we're analyzing, as defined by [Spotify](https://developer.spotify.com/documentation/web-api/reference/tracks/get-audio-features/) (summaries of definitions below)
- **danceability:** how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity  
- **energy:** a perceptual measure of intensity and activity  
- **key:** estimated overall key of the track  
- **loudness:** overall loudness of a track in decibels  
- **mode:** modality (major or minor) of a track  
- **speechiness:** the presence of spoken words in a track  
- **acousticness:** a confidence measure of whether the music is acoustic  
- **instrumentalness:** predicts whether a track contains no vocals  
- **liveness:** detects the presence of an audience in the recording  
- **valence:** the musical positiveness conveyed by a track  
- **tempo:** overall estimated tempo of a track in beats per minute  

It looks like there is a particular interesting relationship between time and acousitcness, danceability, loudness, and valence!
```{r overall, warn = FALSE, message = FALSE, fig.height = 5, fig.width = 10}
plot_metrics(playlist_data, metrics)
```

It looks like there is a particular interesting relationship between time and acousticness, danceability, loudness, and valence! Let's look at these more closely.

### Acousticness
```{r acousticness, echo = FALSE, warn = FALSE, message = FALSE}
plot_metric(playlist_data, "acousticness")
```
Music has gotten *less* acousticness since the '50s', but with an increased bump in acousticness in the '10s (and a big increase in variance) it might be making a comeback.

Here are the five most "acoustic" songs from '50s:
```{r acoustic50s, echo = FALSE}
find_tracks(playlist_data, "All Out 50s", "acousticness", 5) %>%
  knitr::kable
```
...and the top five "acoustic" songs in the '10s that may be driving a comeback:

```{r acoustic10s, echo = FALSE}
find_tracks(playlist_data, "All Out 10s", "acousticness", 5) %>%
  knitr::kable
```

### Danceability
```{r danceability, warn = FALSE, message = FALSE}
plot_metric(playlist_data, "danceability")
```
Though not totally linearly, music has generally become more "dancey" since the 50s, reaching a max in the 2000s. 

These are the top classics that made the 2000s so dancey :notes:
```{r, echo = FALSE}
find_tracks(playlist_data, "All Out 00s", "danceability") %>%
  knitr::kable()
```

### Loudness :microphone:
Music's gotten louder! With a bump in the '10s. Here is the loudest song in each decades playlist:
```{r, echo = FALSE}
pdf <- lapply(names(playlist_data), function(x) {
  find_tracks(playlist_data, x, "loudness", 5) %>%
    mutate(decade = x)
}) %>% bind_rows() %>%
  # remove outliers
  filter(Loudness > -60, Loudness < 0) %>%
  group_by(decade) %>%
  filter(Loudness == max(Loudness)) %>%
  ungroup() %>%
  mutate(decade = factor(decade, levels = paste0("All Out ", years, "s"))) 
pdf %>%
  ggplot(aes(x = decade, y = Loudness)) +
  geom_bar(stat = "identity", aes(fill = Loudness)) +
  geom_text(aes(label = str_wrap(Name, 10)), y = 1, fontface = "bold", 
            size = 3.5, lineheight = .8, vjust = 0) +
  geom_text(aes(label = str_wrap(Artist, 10)), y = 0.75, size = 3, lineheight = .8, vjust = 1) +
  scale_y_continuous(limits = c(-6, 2.5)) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "bottom") + 
  labs(x = "Decade", title = "Loudest Song in Each Decade Playlist",
       y = "Loudness (dBs)")
```

### Valence :blush:
```{r valence, warn = FALSE, message = FALSE}
plot_metric(playlist_data, "valence")
```

Music has been getting sadder! Here are the songs with the highest (happiest) and lowest (saddest) valence in each of the decades playlists:
```{r}
happy <- lapply(names(playlist_data), function(x) {
  find_tracks(playlist_data, x, "valence", 1) %>%
    mutate(decade = x)
}) %>% bind_rows() 

sad <- lapply(names(playlist_data), function(x) {
  find_tracks(playlist_data, x, "valence", 1, desc = FALSE) %>%
    mutate(decade = x)
}) %>% bind_rows() 

line_breaks <- function(nvar) {
  vapply(seq_along(nvar), function(i) {
    paste0(rep("\n", nvar[i] + 2), collapse = "")
  }, character(1))
}

plot_df <- bind_rows(mutate(happy, metric = "happy"),
          mutate(sad, metric = "sad")) %>%
  mutate(decade = factor(decade, levels = paste0("All Out ", years, "s")),
         Name = str_wrap(Name, 15),
         lines = str_count(Name, "\n")) 
plot_df$Artist <- paste0(line_breaks(plot_df$lines), plot_df$Artist)

grouped_df <- plot_df %>% 
  select(decade, metric, Valence) %>% 
  tidyr::spread(metric, Valence)
plot_df %>%
  ggplot(aes(x = decade, y = Valence)) +
  geom_segment(data = grouped_df, aes(xend = decade, y = happy, yend = sad),
               size = 10, color = "grey50", alpha = .25) +
  geom_point(aes(color = metric), size = 10) +
  scale_color_manual(values = c("#ED7B68", "#1E5260"),
                     labels = c("Highest Valence", "Lowest Valence"), 
                     name = NULL) +
  coord_flip() +
  geom_text(
    aes(label = Name,
        hjust = ifelse(metric == "happy", 1, 0),
        y = Valence + ifelse(metric == "happy", -1, 1) * .03
        ), size = 3.5,
    lineheight = .8, fontface = "bold"
  )+
  geom_text(
    aes(label = Artist,
        hjust = ifelse(metric == "happy", 1, 0),
        y = Valence + ifelse(metric == "happy", -1, 1) * .05
        ), size = 3.5,
    lineheight = .8
  ) +
  theme(panel.grid.major.y = element_blank(),
        legend.position = "bottom")
```

Let's look at how they've been distributed over time...
```{r, echo = FALSE}
playlist_data %>% 
  bind_rows() %>% 
    mutate(name = factor(name, levels = paste0("All Out ", years, "s"))) %>%
  ggplot(aes(x = valence, color = name)) + 
  geom_density(size = 2) + 
  scale_color_manual(
    values = grDevices::colorRampPalette(c("#b3e6ff", "#00111a"))(8), name = "Playlist") +
  labs(title = "Distribution of Valence Over Time",
       x = "Valence", y = "Density")
```